Synthesizing Symmetric Lenses

### 2019-02-27 papers/SymmetricLenses/main.tex
make 25 pages yay

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages, like Boomerang, are excellent targets for search-based synthesis, since their
specialized, sub-Turing-complete syntax and expressive type systems drastically
limit the space of possible programs.
%
In particular, the recent Optician~\cite{optician} enhancement to Boomerang has
demonstrated that it is feasible to synthesize quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Specifically, given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists. Indeed, it can often find
an appropriate lens in less than a second with no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

@@

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly the symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  To check that they are expressive
enough in practice, we add simple symmetric lenses to the Boomerang
language~\cite{boomerang} and apply them to a range of real-world
applications, showing in the process that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

@@

We consider a ``likely'' lens to be one that propagates ``a lot'' of information
from the left data format to the right and vice versa. Conversely, an unlikely
lens is one that requires a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the likelihood of a given lens.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure. The
interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to iteratively propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, takes a pair of SREs as input and uses a greedy, type-directed
algorithm to find a simple symmetric lens between them, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

@@

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing the source and target types and optionally supplying one or
more examples. In addition, users can override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
thereby forcing certain data to either be retained or discarded during the
transformations. We evaluated our system on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).

@@

\item We also extend Boomerang with symmetric lens synthesis specifications,
  which include types, examples, and relevance annotations. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.

@@

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other. One manifestation of this principle involves choosing the
identity lens (which uses all the data) rather than the disconnect lens (which
uses none of the data). We define the likelihood of a satisfying lens as the
expected number of bits required to recover data in one format from synchronized
data in the other. Higher likelihoods correspond to fewer bits. We say that two
strings $s$ and $t$ are \emph{synchronized according to lens $l$} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We say that we can \emph{recover} $s$ from
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, each element of $s \in S$ has \emph{information content},
the number of bits required to specify that element in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ calculates the expected number of bits
required to specify a random element pulled from $S$, in other words, the
probability of each element times its information content. Entropy maintains
that if a data source contains many possible elements, where none have
significantly higher probability than others, it will have high information
content. Data sources with few elements that don't have vanishingly small
probabilities have small entropy. When $P$ is clear, we will often use the
shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we must first develop a probabilistic 
model for our language. We do this with \emph{stochastic regular
  expressions} (SREs)---regular expressions annotated with probability
information~\cite{stoch-rnn,stoch-def}---which 
jointly express a language and a probability distribution over that language. 
$$\Regex{},\RegexAlt{} \quad\GEq{}\quad  \String  \quad|\quad \emptyset \quad|\quad \PRegexStar{\Regex}{p} \quad|\quad \RegexConcat{\Regex_1}{\Regex_2} \quad|\quad \PRegexOr{\Regex_1}{\Regex_2}{p}$$
(Lowercase $\String$ ranges over constant strings and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.)

The semantics of an SRE $S$ is a probability distribution $P_S$ defined as
follows.
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\String}{\String''}$
    & =
    & $\begin{cases*}1 & if $\String = \String''$\\ 0 & otherwise\end{cases*}$ \\

@@

    $\ProbabilityOf{\PRegexStar{\Regex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n(1-\Probability)\Pi_{i=1}^n\ProbabilityOf{\Regex}{\String_i}$\\
  \end{tabular}
\end{center}
In the SRE $\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$, the annotation
$\Probability$ represent that, if generating arbitrary strings from
$\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$, strings should be pulled from
$\Regex_1$ with probability $\Probability$ (\EG pulling a string from
$\PRegexOr{\lstinline{"a"}}{\lstinline{"b"}}{.2}$ would retrieve \lstinline{"a"}
20\% of the time). In the SRE
$\PRegexStar{\Regex}{\Probability}$, the annotation \Probability represents how
much more likely having a string involving more than $n$ iterations is than
having a string involving exactly $n$ iterations (\EG pulling a random string
from $\PRegexStar{\lstinline{"c"}}{.8}$ would give \lstinline{""} 20\% of the
time, \lstinline{"c"} 16\% of the time, and \lstinline{"cc"} 12.8\% of the time).

@@

\subsection{Stochastic Regular Expression Equivalences}
The $\Expand$ algorithm needs to be able to enumerate SREs that are
``equivalent'' to a given one. However, existing work does not define any notion
of stochastic regular expression equivalence. Figure~\ref{fig:ssr-sre} shows how
we to extend the \emph{star-semiring}~\cite{Droste2009} equivalences (a
\emph{finer} notion of equivalence than semantic equivalence) to
SREs.

@@

\subsection{Stochastic Regular Expression Entropy}
The entropy of a data source $S$ is the expected number of bits required to
describe a random element in $S$ (formally $-\Sigma_{s\in S}\,P(s)\cdot
\Log_2P(s)$). The entropy of a SRE can be computed directly from its syntax,
when each string is uniquely parsable (\IE{} the SRE is unambiguous) and contains
no empty subcomponents.
\begin{center}
  \begin{tabular}{rcl}
    $\EntropyOf{s}$
    & =
    & 0\\

@@

For example, the entropy of $\PRegexOr{\lstinline{"a"}}{(\lstinline{"b"})}{.5}$
is 1. This is expected because on average 1 bit is required to encode whether a
randomly pulled string is \lstinline{"a"} or \lstinline{"b"}. As an additional
example, a fixed string has no information content, and so has no entropy.

@@

The procedure \PCF{\Continue} returns false based on the how long the search has
been going, and how hard it expects the next class of problems to be. In
particular, if $\PCF{PQ.Peek}(pq) = (\Regex,\RegexAlt)$, that RE pair is at
distance $d$, the number of pairs in $pq$ at distance $d$ is $n$, and the
current best lens has cost $c$, then $\Call{Continue}{pq}$ checks whether $c > d
+ \Log_2(n)$. This termination condition is based around two primary principles:
do not search overly deep ($d$), and do not tackle a frontier that would take
too long to process ($\Log_2(n)$). The log of the frontier is included because
the frontier grows much faster than lens costs grow. Lens cost grows with the
log of the expected number of choices in a given lens (due to its information
theoretic basis), so the frontier calculation does too.

When $\Call{Continue}{pq}$ returns $\False$, the algorithm stops proposing
regular expression pairs, and instead returns to the user the best lens found
thus far. If the algorithm finds a bijective lens, which
has zero cost, it will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
\label{subsec:sdnfre}
The \GreedySynth subroutine converts input SREs into a temporary pseudo-normal
form, \emph{Stochastic DNF regular expressions} (\SDNFREabbrev{}s).
\SDNFREabbrev{}s normalize across many of the star-semiring equivalences -- if
two regular expressions are equal modulo differences in associativity,
commutativity, or distributivity, their corresponding SREs are syntactically
equal.

@@

The algorithm for converting a stochastic regular expressions into its
corresponding stochastic DNF regular expressions, written
$\ToDNFRegexOf{\Regex}$, is defined below. This conversion relies on operators
defined in Figure~\ref{fig:dnf-regex-functions}.
\[
  \begin{array}{rcl@{\hspace*{3em}}rcl}
    \ToDNFRegexOf{\String} 
    & = 
    & \DNFOf{(\SequenceOf{\String},1)}

@@

  \end{array}
\]
After this conversion has taken place, the sequences are ordered (thereby
normalizing for commutativity differences).
We have proven that this conversion respects languages and probability
distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain \ToStochastic, the function that
converts regular expressions into stochastic regular expressions. If $\Regex$ is
a stochastic regular expression generated by \ToStochastic, then when put into
\SDNFREabbrev form,
$\ToDNFRegexOf{\Regex} = \DNFOf{(\Sequence_1,\frac{1}{n}) \DNFSep \ldots \DNFSep
  (\Sequence_n,\frac{1}{n})}$ for some sequences $\Sequence_1\ldots\Sequence_n$,
and every stochastic atom generated by \ToStochastic is
$\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$ generates regular
expressions whose DNF form gives equal probability to all sequence subcomponents
of the DNF regular expression, and gives iterations a .8 chance to continue
iterating. In our experience generating random strings from regular expressions,
these probabilities provide good distributions of strings---stars are iterated 4
times on average, and no individual choice in a series of disjunctions is chosen
disproportionately often.

@@

Even though our generated probability distribution works well in most
situations, it is not perfect. Consider synthesizing a lens between the formats
shown in Figure~\ref{fig:minimized-representations}. While salary information is
present in \lstinline{emp_salaries}, but not in \lstinline{emp_insurance}, the
algorithm might spend a long time (fruitlessly) trying to construct lenses that
transform the salary to information present in \lstinline{emp_insurance}.

One solution would be to cut off the search early. However, then one runs into
the opposite problem: in other scenarios salary information may be present in
the other format, but it may well take quite a bit of work to find a
transformation that connects the salary in \lstinline{emp_salaries} to the
salary information in the second format. In such a scenario, early termination
may leave salary information discarded where it should be retained.

@@

For example, the user can easily recognize that salary information of
\lstinline{emp_salaries} is not present in \lstinline{emp_insurance}. By
annotating the \lstinline{salary} field as \lstinline{skip(salary)}, users can
add this knowledge to the specification to optimize the search. In practice,
we define \Skip by setting the information content of $\Skip(\Regex)$ to zero.
\begin{gather*}
  \EntropyOf{\Skip(\Regex)} = 0
\end{gather*}

Similarly, users can recognize that employee names are present in both files. By
annotating instances of \lstinline{name} as \lstinline{require(name)}, users can
add this knowledge to the specification to optimize the search and enforce that
the returned lens that retains \lstinline{name} information. In practice, we
define \SRequire by setting the cost of any lens that loses ``required''
information to infinity.
%
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}$
    & =
    & $\begin{cases}
      \infty & \text{if }\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}
      \neq 0\\
    0 & \text{otherwise}
  \end{cases}$\\

@@

The structure of symmetric DNF lenses exactly mirrors that of SDNF REs. A
symmetric DNF lens $sdl$ is a union of symmetric sequence lenses, $ssql_1 \ldots
ssql_p$, a symmetric DNF sequence lens is a concatenation of symmetric atom
lenses, $sal_1 \ldots sal_n$, and a symmetric atom lens $sal$ is an iteration of
a symmetric DNF lens.

@@

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lens is popped from the queue and, if it is
``useful,'' it is included in the final DNF lens. The lens is considered to
be useful when its source (or target) is \textit{not} already the source (or target) of
an already chosen sequence lens. If the examples require that two sequences have a
lens between them, then any sequence lens between them is considered useful. We
update the priorities of the sequence lenses as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had, to account for the information that needs to be stored
for including that source of non-bijectivity.

@@

Our implementation includes a number of optimizations not described above:
annotations that guide DNF conversion; an expansion inference algorithm;
hash-consing and memoization; and compositional synthesis. The optimizations
make the system performant enough for interactive use, and the annotations help
ensure the desired information is projected.

@@

\paragraph*{Hash-consing and Memoization} We further improve \GreedySynth by
hash-consing stochastic DNF regular expressions.  This optimization enables
fast equality checks and saves memory.
%memoizing its results.  However, our stochastic DNF regular expressions are
%quite large and so hashing them is computationally intensive.  We use less
%memory and allow for quick comparisons and hashing of these DNF regular
%expressions by

@@

Because our tool is intended to be interactive---the user typically starts with
zero or one examples, runs the tool, determines whether the output is the
desired one and, if not, either adds an example or relevance annotations and
then tries again---it is hard to cleanly evaluate the net user experience. The
primary concern is whether it is usable to synthesize correct lenses.

>>>>

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages, like Boomerang, are excellent targets for search-based synthesis, since their
specialized, sub-Turing-complete syntax and expressive type systems drastically
limit the space of possible programs.
%
The recent Optician~\cite{optician} enhancement to Boomerang has
demonstrated that it is feasible to synthesize quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists. Indeed, it can often find
an appropriate lens in less than a second with no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection, it is fairly likely to be the intended one.

@@

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  We show they are expressive
enough in practice by adding simple symmetric lenses to the Boomerang
language~\cite{boomerang} and applying them to a range of real-world
applications, showing that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

@@

We consider ``likely'' lenses to be ones that propagate ``a lot'' of information
from the left data format to the right and vice versa. Conversely, unlikely
lenses require a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the likelihood of a given lens.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure.
Interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types to find the most likely lens. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, uses a greedy, type-directed
algorithm to find a simple symmetric lens between input SRE pairs, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when a sufficiently likely lens is found.

@@

We added this synthesis procedure to the Boomerang system and explored its
effectiveness on a range of applications. Users set up a Boomerang synthesis
task by providing the source and target types and optionally supplying one or
more examples. In addition, users can override the default mechanism for
calculating the information content of a SRE by
asserting that certain strings are \emph{essential} or \emph{irrelevant},
thereby forcing certain data to either be retained or discarded during the
transformations. We evaluated our system on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{flashfill}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).

@@

\item We also extend Boomerang with symmetric lens synthesis specifications:
  including types, examples, and relevance annotations. We evaluate our implementation, the effects of
  optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show the {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices
\ifappendices\else included with the supplemental material\fi.

@@

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it uses more data from one format to
construct the other. One manifestation of this principle involves choosing the
identity lens (which uses all the data) rather than the disconnect lens (which
uses none of the data). We define the likelihood of a satisfying lens as the
expected number of bits required to recover data in one format from synchronized
data in the other. Higher likelihoods correspond to fewer bits. Two
strings $s$ and $t$ are \emph{synchronized according to lens $l$} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We can \emph{recover} $s$ from
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, each element of $s \in S$ has \emph{information content},
the number of bits required to specify that element in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ calculates the expected number of bits
required to specify an element drawn from $S$, in other words, the
probability of each element times its information content. Entropy maintains
that if a data source contains many possible elements, where none have
significantly higher probability than others, it will have high information
content. Data sources with few elements that don't have vanishingly small
probabilities have small entropy. When $P$ is clear, we will often use the
shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we first develop a probabilistic 
model for our language using \emph{stochastic regular
  expressions} (SREs)---regular expressions annotated with probability
information~\cite{stoch-rnn,stoch-def}---that
jointly express a language and a probability distribution over that language. 
$$\Regex{},\RegexAlt{} \quad\GEq{}\quad  \String  \quad|\quad \emptyset \quad|\quad \PRegexStar{\Regex}{p} \quad|\quad \RegexConcat{\Regex_1}{\Regex_2} \quad|\quad \PRegexOr{\Regex_1}{\Regex_2}{p}$$
(Lowercase $\String$ ranges over constant strings and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.)
The semantics of an SRE $S$ is a probability distribution $P_S$ defined as
follows.
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\String}{\String''}$
    & =
    & $\begin{cases*}1 & if $\String = \String''$\\ 0 & otherwise\end{cases*}$ \\

@@

    $\ProbabilityOf{\PRegexStar{\Regex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n(1-\Probability)\Pi_{i=1}^n\ProbabilityOf{\Regex}{\String_i}$\\
  \end{tabular}
\end{center}
One may think of SREs as string generators. Under this interpretation, the
constant SRE $\String$ always generates the string $\String$ and never any other
string. The SRE $\PRegexOr{\Regex_1}{\Regex_2}{\Probability}$ generates a string
from $\Regex_1$ with probability $\Probability$ and generates a string from
$\Regex_2$ with probability $1-\Probability$. The SRE
$\PRegexStar{\lstinline{"c"}}{\Probability}$ generates the empty string with
probability $1-\Probability$, and generates any other string in the language with
probability $\Probability$. More generally,
$\PRegexStar{\Regex}{\Probability}$ generates strings in $\Regex^n$ with
probability $\Probability^n(1-\Probability)$.

@@

\subsection{Stochastic Regular Expression Equivalences}
The $\Expand$ algorithm enumerates SREs that are
``equivalent'' to a given one. However, existing work does not define any notion
of stochastic regular expression equivalence. Figure~\ref{fig:ssr-sre} shows how
we to extend the \emph{star-semiring}~\cite{starsemiring} equivalences (a
\emph{finer} notion of equivalence than semantic equivalence) to
SREs.

@@

\subsection{Stochastic Regular Expression Entropy}
The entropy of a data source $S$ is the expected number of bits required to
describe an element drawn from $S$ (formally $-\Sigma_{s\in S}\,P(s)\cdot
\Log_2P(s)$). The entropy of a SRE can be computed directly from its syntax,
when each string is uniquely parsable (\IE{} the SRE is unambiguous) and contains
no empty subcomponents.
\begin{center}
  \begin{tabular}{rcl}
    $\EntropyOf{s}$
    & =
    & 0\\

@@

For example, the entropy of $\Regex =
\PRegexOr{\lstinline{"a"}}{(\lstinline{"b"})}{.5}$ is 1. The best encoding of a
stream of elements from $\Regex$ will use, on average, 1 bit per element to
determine whether that element is \lstinline{"a"} or \lstinline{"b"}. As an
additional example, a fixed string has no information content, and so has no
entropy.

@@

The procedure \PCF{\Continue} terminates the loop based on the how long the
search has been going, and how hard it expects the next class of problems to be.
In particular, if $\PCF{PQ.Peek}(pq) = (\Regex,\RegexAlt)$, that RE pair is at
distance $d$, the number of pairs in $pq$ at distance $d$ is $n$, and the
current best lens has cost $c$, then $\Call{Continue}{pq}$ continues the loop
while $c < d + \Log_2(n)$. This termination condition is based around two
primary principles: do not search overly deep ($d$), and do not tackle a
frontier that would take too long to process ($\Log_2(n)$). The log of the
frontier is included because the frontier grows much faster than lens costs
grow. Lens cost grows with the log of the expected number of choices in a given
lens (due to its information theoretic basis), so the frontier calculation does
too.

When $\Call{Continue}{pq}$ terminates the loop, the algorithm stops proposing
regular expression pairs, and instead returns to the user the best lens found
thus far. If the algorithm finds a bijective lens, which
has zero cost, it will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
\label{subsec:sdnfre}
The \GreedySynth subroutine converts input SREs into a temporary pseudo-normal
form, \emph{Stochastic DNF regular expressions} (\SDNFREabbrev{}s).
\SDNFREabbrev{}s normalize across many of the star-semiring equivalences -- if
two regular expressions are equal modulo differences in associativity,
commutativity, or distributivity, their corresponding SREs are syntactically
equal.

@@

The algorithm for converting a stochastic regular expressions $\Regex$ into its
corresponding \SDNFREabbrev form, written
$\ToDNFRegexOf{\Regex}$, is defined below. This conversion relies on operators
defined in Figure~\ref{fig:dnf-regex-functions}.
\[
  \begin{array}{rcl@{\hspace*{3em}}rcl}
    \ToDNFRegexOf{\String} 
    & = 
    & \DNFOf{(\SequenceOf{\String},1)}

@@

  \end{array}
\]
After this syntactic conversion has taken place, the sequences are ordered
(normalizing commutativity differences). This conversion respects languages and
probability distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain \ToStochastic, the function that
converts regular expressions into stochastic regular expressions. If $\Regex$ is
a stochastic regular expression generated by \ToStochastic, then when put into
\SDNFREabbrev form, $\ToDNFRegexOf{\Regex} = \DNFOf{(\Sequence_1,\frac{1}{n})
  \DNFSep \ldots \DNFSep (\Sequence_n,\frac{1}{n})}$ for some sequences
$\Sequence_1\ldots\Sequence_n$, and every stochastic atom generated by
\ToStochastic is $\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$
generates regular expressions whose DNF form gives equal probability to all
sequence subcomponents of the \SDNFREabbrevs, and gives a .8 chance for
stars to continue iterating. In our experience generating random strings from regular
expressions, these probabilities provide good distributions of strings---stars
are iterated 4 times on average, and no individual choice in a series of
disjunctions is chosen disproportionately often.

@@

Even though our generated probability distribution works well in most
situations, it is not perfect. Consider synthesizing a lens between the formats
shown in Figure~\ref{fig:minimized-representations}. Because salary information is
present in \lstinline{emp_salaries}, but not in \lstinline{emp_insurance}, the
algorithm might spend a long time (fruitlessly) trying to construct lenses that
transform the salary to information present in \lstinline{emp_insurance} even
though that is impossible. In a similar but more elaborate example, such wasted
processing effort may cause synthesis to fail to terminate in any reasonable
amount of time.

One solution would be to cut off the search early. However, then one runs into
the opposite problem: In other scenarios, salary information may be present in
the other format, but it may take quite a bit of work to find a
transformation that connects the salary in \lstinline{emp_salaries} to the
salary in the second format. Hence, early termination
may cut off synthesis before the right lens is found.

@@

For example, users can easily recognize that salary information is not present
in \lstinline{emp_insurance}. By annotating the \lstinline{salary} field as
\lstinline{skip(salary)}, users can add this knowledge to the specification to
optimize the search. In practice, we define the information content of
$\Skip(\Regex)$ to be zero.
\begin{gather*}
  \EntropyOf{\Skip(\Regex)} = 0
\end{gather*}

Similarly, users can recognize that employee names are present in both files. By
annotating instances of \lstinline{name} as \lstinline{require(name)}, users can
add this knowledge to the specification to optimize the search and force the
the generated lens to retain \lstinline{name} information. In practice, we
define the cost of any lens that loses ``required'' information to be infinite.
%
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}$
    & =
    & $\begin{cases}
      \infty & \text{if }\REntropyOf{\SRequireOf{\RegexAlt} \Given \Lens, \Regex}
      \neq 0\\
    0 & \text{otherwise}
  \end{cases}$\\

@@

The structure of symmetric DNF lenses mirrors that of SDNF REs. A symmetric DNF
lens ($sdl$) is a union of symmetric sequence lenses, a symmetric sequence
lens ($ssql$) is a concatenation of symmetric atom lenses, and a
symmetric atom lens ($sal$) is an iteration of a symmetric DNF lens.

@@

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lens is popped from the queue and, if it is
``useful,'' is included in the final DNF lens. The lens is considered to be
useful when its source (or target) is \textit{not} already the source (or
target) of an included sequence lens. If examples require that two
sequences have a lens between them, such lenses are considered useful. The
priorities of the sequence lenses update as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had; information must now be stored for including that
source of non-bijectivity.

@@

Our implementation includes a number of optimizations not described above:
annotations that guide DNF conversion; an expansion inference algorithm;
and compositional synthesis. The optimizations
make the system performant enough for interactive use, and the annotations help
ensure the desired information is projected.

@@

@@
